### 2011-02-09 paper/hyperNeatMethod.tex
hyperneat edits

\emph{Neuroevolution (Evolving Artificial Neural Networks with HyperNEAT)}:

HyperNEAT is an indirect encoding for evolving artificial neural
networks (ANNs) that is inspired by the way natural organisms
develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
which is a genome that encodes an ANN
phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
graph, where each node is a mathematical function, such as sine or
Gaussian. The nature of these functions can facilitate the evolution
of properties such as symmetry (e.g., an absolute value or Gaussian
function) and repetition (e.g., a sine
function)~\cite{stanley2009hypercube, stanley2007compositional}. The
signal on each link in the CPPN is multiplied by that link's weight,
which can alter its effect.

A CPPN is queried once for each link in the ANN substrate to determine
that link's weight. The inputs to the CPPN are the Cartesian
coordinates of both the source (e.g., $x = 2$, $y = 4$) and target
(e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias
value. The CPPN takes these five values as inputs and produces one or
two output values, depending on the substrate topology. If there is no
hidden layer in the substrate, the single output is the weight of the
link between a source node on the input layer and a target node on the
output layer. If there is a hidden layer, the first output value
determines the weight of the link between the associated input
(source) and hidden layer (target) nodes, and the second output value
determines the weight of the link between the associated hidden
(source) and output (target) layer nodes. All pairwise combinations of
source and target nodes are iteratively passed as inputs to a CPPN to
determine the weight of each substrate link.

@@

HyperNEAT is capable of exploiting the geometry of a problem: because
the link values between nodes in the final ANN substrate are a
function of the geometric positions of those nodes, HyperNEAT can
exploit such information when solving a
problem~\cite{stanley2009hypercube, clune2009sensitivity,
  clune2011performance}. In the case of quadruped locomotion, this
property helped Hyper-NEAT produce gaits with front-back, left-right,
and four-way symmetries~\cite{clune2009evolving,
  clune2011performance}.

The evolution of the population of CPPNs occurs according to the
principles of the NeuroEvolution of Augmenting Topologies (NEAT)
algorithm~\cite{stanley2002evolving}, which was originally designed to
evolve ANNs. NEAT can be fruitfully applied to CPPNs because of their
structural similarity to ANNs. For example, mutations can add a node,
and thus a function, to a CPPN graph, or change its link weights. The
NEAT algorithm is unique in three main
ways~\cite{stanley2002evolving}. Initially, it starts with small
genomes that encode simple networks and slowly complexifies them via
mutations that add nodes and links to the network, enabling the
algorithm to evolve the network topology in addition to its
weights. Secondly, NEAT has a fitness sharing mechanism that preserves
diversity in the system and gives time for new innovations to be tuned
by evolution before competing them against more adapted
rivals. Finally, NEAT tracks historical information to perform
intelligent crossover while avoiding the need for expensive
topological analysis. A full explanation of NEAT can be found
in~\cite{stanley2002evolving}.

The ANN configuration follows previous studies that evolved quadruped
gaits with HyperNEAT in simulation~\cite{clune2011performance,
  clune2009evolving}, but was adapted to accommodate differences in
the physical robot. Specifically, the ANNs consists of three $3 \times 4$
Cartesian grids of nodes forming input, hidden, and output
layers. Adjacent layers were completely connected, meaning that there
were $(3 \times 4)^2= 288$ links in each substrate. The inputs to the
substrate were the current angles of each of the 9 joints of the robot
(described below) and a sine and cosine wave (to facilitate the
production of periodic behaviors). The sine and cosine waves were the
following function of time ($t$) in milli-seconds:
$\sin(t/12)\pi$. This function produces numbers in the range [$-\pi,
  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
COMMANDS WERE SPECIFIED.

@@

\acmFig{SpiderANN.jpg}{1}{ANN Configuration for HyperNEAT Runs. The
  first two columns of each row of the input layer receive information
  about a single leg (the current angle of its two joints). The final
  column provides a sine and cosine wave to enable periodic movements
  and the angle of the center joint. Evolution determines the function
  of the hidden-layer nodes. The nodes in the output layer specify new
  joint angles for each respective joint. The unlabeled node in the
  input and output layer is ignored.}

The population size for HyperNEAT was 9 and runs lasted 20
generations. These numbers are small, but were necessarily constrained
given how much time it took to conduct evolution directly on a real
robot. The remaining parameters were identical to Clune et
al. ~\cite{clune2011performance}.

>>>>

HyperNEAT is an indirect encoding for evolving artificial neural
networks (ANNs) that is inspired by the way natural organisms
develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
which is a genome that encodes an ANN
phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
graph, where the nodes in the graph are mathematical functions, such as sine or
Gaussian. The nature of these functions can facilitate the evolution
of properties such as symmetry (e.g.\ a Gaussian function) and repetition (e.g.\ a sine
function)~\cite{stanley2009hypercube, stanley2007compositional}. The
signal on each link in the CPPN is multiplied by that link's weight,
which can magnify or diminish its effect.

A CPPN is queried once for each link in the ANN phenotype to determine
that link's weight. The inputs to the CPPN are the Cartesian
coordinates of both the source (e.g.\ $x = 2$, $y = 4$) and target
(e.g.\ $x = 3$, $y = 5$) nodes of a link and a constant bias
value. The CPPN takes these five values as inputs and produces two output values. The first output value
determines the weight of the link between the associated input
(source) and hidden layer (target) nodes, and the second output value
determines the weight of the link between the associated hidden
(source) and output (target) layer nodes. All pairwise combinations of
source and target nodes are iteratively passed as inputs to a CPPN to
determine the weight of each ANN link.

@@

HyperNEAT is capable of exploiting the geometry of a problem because
the link values between nodes in the final ANN phenotype are a
function of the geometric positions of those nodes~\cite{stanley2009hypercube, clune2009sensitivity,
  clune2011performance}. In the case of quadruped locomotion, this
property has previously been shown to help Hyper-NEAT produce gaits with front-back, left-right,
and four-way symmetries~\cite{clune2009evolving,
  clune2011performance}.

The evolution of the population of CPPNs occurs according to the
principles of the NeuroEvolution of Augmenting Topologies (NEAT)
algorithm~\cite{stanley2002evolving}, which was originally designed to
evolve ANNs. NEAT can be fruitfully applied to CPPNs because of their
structural similarity to ANNs. For example, mutations can add a node,
and thus a function, to a CPPN graph, or change its link weights. The
NEAT algorithm is unique in three main
ways~\cite{stanley2002evolving}. Initially, it starts with small
genomes that encode simple networks and slowly complexifies them via
mutations that add nodes and links to the network, enabling the
algorithm to evolve the topology of an ANN in addition to its
weights. Secondly, NEAT has a fitness-sharing mechanism that preserves
diversity in the system and gives time for new innovations to be tuned
by evolution before competing them against more adapted
rivals. Finally, NEAT tracks historical information to perform
intelligent crossover while avoiding the need for expensive
topological analysis. A full explanation of NEAT can be found
in~\cite{stanley2002evolving}.

The ANN configuration follows previous studies that evolved quadruped
gaits with HyperNEAT in simulation~\cite{clune2011performance,
  clune2009evolving}, but was adapted to accommodate the physical robot used in this paper. Specifically, the ANNs consists of three $3 \times 4$
Cartesian grids of nodes forming input, hidden, and output
layers (\figref{SpiderANN.jpg}). Adjacent layers were allowed to be completely connected, meaning that there
could be $(3 \times 4)^2= 288$ links in each ANN (although evolution can set weights to 0, functionally eliminating the connection). The inputs to the
substrate were the current angles of each of the 9 joints of the robot
(described below) and a sine and cosine wave (to facilitate the
production of periodic behaviors). The sine and cosine waves were the
following function of time ($t$) in milli-seconds:
$\sin(t/12)\pi$. This function produces numbers in the range [$-\pi,
  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
COMMANDS WERE SPECIFIED.

@@

\acmFig{SpiderANN.jpg}{1}{ANN Configuration for HyperNEAT Runs. The
  first two columns of each row of the input layer receive information
  about a single leg (the current angle of its two joints). The final
  column provides a sine and cosine wave to enable periodic movements
  and the angle of the center joint. Evolution determines the function
  of the hidden-layer nodes. The nodes in the output layer specify new
  joint angles for each respective joint. The unlabeled node in the
  input and output layer is ignored.}\label{hyperneatLayout}

The population size for HyperNEAT was 9 and runs lasted 20
generations. These numbers are small, but were necessarily constrained
given how much time it took to conduct evolution directly on a real
robot. The remaining parameters were identical to Clune et
al.~\cite{clune2011performance}.
